Analysing Constraint Grammars with a SAT-solver

### 2016-02-16 paper/lrec/eval_futurework.tex
more paper

\paragraph{Accuracy} 
The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars.

The Spanish grammar had an erroneous set definition, where a word was
required to have almost all POS tags at the same time. After fixing
the set definition, our test did not reveal other problems.

The Dutch grammar had two kinds of errors: rule-internal due to nonexistent tags, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

@@

For the Finnish grammar, we did not have time to investigate all the
results. We could verify approximately \nicefrac{1}{3} as true positives and
\nicefrac{1}{3} as false positives.
The verified true positives were all rule-internal conflicts. 
For example, the following rule requires the target to be genitive and unambiguously nominative: 
\begin{itemize}
\item[]\texttt{"oma" SELECT GEN IF [...] (0C NOM) ;}
\end{itemize}
We also detected a number of tagset errors;
the grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

@@

Our solution to hardcode the tag combinations in the readings is feasible for simple morphology, but it can cause problems with more complex morphology. One big downside is that in order to implement \textsc{ADD}, \textsc{ADDREADING} and \textsc{MAP}, we need to be prepared for new readings--even if the lexicon gives all readings that exist in the lexicon, the user might give a nonexistent reading, or in the case of MAP, a syntactic tag, which is (by definition) not in the lexicon. A more scalable solution would be to make each tag a variable, and ask the question ``can this reading be a noun? how about singular? how about conditional?'' separately for each tag. Then we could lift the restriction of tag combinations into a SAT solver: make SAT clauses that prohibit a comparative to go with a verb, or conditional with a noun.

>>>>

\paragraph{Accuracy} 
The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars. However, the test reported a high number of errors even in the smallest grammars, and some of them are intuitively not that bad; we

\paragraph{Dutch} The Dutch grammar had two kinds of errors: rule-internal due to nonexistent tags, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

@@

\paragrahp{Spanish} The Spanish grammar had proportionately the highest number of errors.\footnote{TODO://link/to/error.log} \todo{confirm; I assume this is gonna be so, but let's see Fin & Por}.  The grammar we ran is like the one found in the Apertium repository, apart from 2 changes: we fixed some typos (capital O for 0) in order to make it compile, and commented out two rules that used regular expressions, because our solution does not handle them properly (see more on Future work/Implementation).

Again, we can classify the conflicts into internal and interaction. As an example of internal conflict, there are \todo{n} rules that use \texttt{SET Cog = (np cog)}---the problem is that the tag \texttt{cog} does not exist in the Apertium dictionary for Spanish. It is likely that this grammar has been written for an earlier version, where such tag has been in place.

The Spanish grammar has a high number of interaction conflicts, all with the same pattern:

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE (a) IF (-1 x) (1 y) ;
REMOVE (a) IF (-1 x) (1 y) (NOT -1 z);
\end{verbatim}
\end{itemize}

This is another, perhaps counter-intuitive instance of more general rule applying before more specific. Negated context works as well to make a rule stricter. The second rule will never apply, because the first will catch all cases.

In addition, there was a number of definitions that were never used. The program doesn't currently point them out to the user, but that is a trivial addition; just like rule-internal conflicts, it doesn't require any semantic analysis. 
Another error, which was not pointed by the program, but discovered

\paragrahp{Finnish} \todo{Rerun the tests & rewrite this whole section!} For the Finnish grammar, we did not have time to investigate all the
results. We could verify approximately \nicefrac{1}{3} as true positives and
\nicefrac{1}{3} as false positives.
The verified true positives were all rule-internal conflicts. 
For example, the following rule requires the target to be genitive and unambiguously nominative: 
\begin{itemize}
\item[]\texttt{"oma" SELECT GEN IF [...] (0C NOM) ;}
\end{itemize}
We also detected a number of tagset errors;
the grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

@@

Our solution to hardcode the tag combinations in the readings is feasible for simple morphology, but it can cause problems with more complex morphology. One big downside is that in order to implement \textsc{ADD}, \textsc{ADDREADING} and \textsc{MAP}, we need to be prepared for new readings--even if the lexicon gives all readings that exist in the lexicon, the user might give a nonexistent reading, or in the case of MAP, a syntactic tag, which is (by definition) not in the lexicon. A more scalable solution would be to make each tag a variable, and ask the question ``can this reading be a noun? how about singular? how about conditional?'' separately for each tag. Then we could lift the restriction of tag combinations into a SAT solver: make SAT clauses that prohibit a comparative to go with a verb, or conditional with a noun.
Another benefit of this solution is the addition of lemmas and word forms, as well as regular expressions: currently, if we want to add one more lemma for a verb, we need to create as many new variables as there are distinct verb forms---easily hundreds for languages with rich morphology.
\todo{less rambling: For rules with regular expressions, this would blow up even more: there are rules that only address whether the word form in a condition starts with a lowercase letter, or whether it ends in a certain suffix. A realistic, albeit unsatisfactory option is just to treat it as an underspecified reading, and offer that same regular expression to the user when generating a symbolic sentence. Another, less feasible, solution is to grep all the word forms or lemmas that match the regular expression in question---in the worst case (e.g. lower case), this will match all the entries in the expanded lexicon.}

### 2016-02-16 paper/lrec/implementation.tex
more paper

 However, for languages with more readings, this may not be feasible. For instance, the Spanish lexicon has a high number of readings mostly because of the verbs: in addition to the verb forms, there are clitics attached to them. The number of verb readings without clitics is 213, and with clitics 1572.

>>>>

 However, for languages with more readings, this may not be feasible. For instance, Spanish has a high number of readings, not only because of many inflectional forms, but because it is possible to add 1-2 clitics to
  lexicon has a high number of readings mostly because of the verbs: in addition to the verb forms, there are clitics attached to them. The number of verb readings without clitics is 213, and with clitics 1572.
